---
title: "RR_Week4_Assignment_on_StormData"
author: "Bishnu Poudel"
date: "March 21, 2019"
output: html_document
---

### Including the preprocessing steps as commented R Scripts

## Preprocessing: Download the file , read the csv, then dump un-necessary columns

```{r   message=FALSE, warning=FALSE}
# #Download file -- Uncomments
# setwd("../DATA_SCIENCE_COURSERA/Reproducible Research/Week4/")
# download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
#               , destfile="Stormdata.csv.dz2", method="libcurl")
# data<- read.csv("Stormdata.csv")
# 
# #Strip out unnecessay columns and write to to different .csv if needed
# #I've under 2GB ram in my system.
# names_loc_out<- c("COUNTY","COUNTYNAME","STATE","LATITUDE","LONGITUDE","LATITUDE_E","LONGITUDE_","REFNUM")
# names_rem_out<-  c( "STATEOFFIC","ZONENAMES", "REMARKS","REFNUM" )
# names_loc_del<- c("COUNTYNAME","STATE","LATITUDE","LONGITUDE"
#                   ,"LATITUDE_E","LONGITUDE_", "STATEOFFIC","ZONENAMES", "REMARKS")
# addressdim<- data[names_loc_out]
# remarks<-data[names_rem_out]
# df<- data[, !(colnames(data) %in% names_loc_del)  ]
# write.csv(addressdim, file="AddressDim.csv")
# write.csv(remarks, file="Remarks.csv")
# rm(data)
```

From findings of initial EDA, the following columns give the crux of the information for addressing our questions on 1. The event with most fatalities/injuries 2. The event with most economic consequences.
1. FATALITIES
2. INJURIES 
3. PROPDMG , PROPDMGEXP (k,m,b,h)
4. CROPDMG,CROPDMGEXP(k,m,b,h)

### Adding one of the EDA done, Finding the granularity of the data
#### First , we investigate the duplicate on  BGN_DATE, COUNTY, EVTYPE

```{r   message=FALSE, warning=FALSE}
# Task 1 - Find what the database is unique on (i.e. primary Key)
# Also exploring data on my Own

#Comment this read part if the above chunk is not run
df<- read.csv("stripped.csv")
dim(df)  # [1] 902297 28
# dim( unique( df[c("BGN_DATE","COUNTY","EVTYPE")] ) )   ##[1] 570401  3
# dim( unique( df[c("BGN_DATE","BGN_TIME", "COUNTY","EVTYPE")] )) ##871125 4
# dim( unique( df[c( "STATEID", "BGN_DATE","BGN_TIME", "COUNTY","EVTYPE")] ) ) ##[1] 873837 5
library(sqldf)
#duplicates on 3 variables
dups<- sqldf(" select BGN_DATE, COUNTY, EVTYPE, count(*) as counts from df group by 1,2,3 having count(*) >1 ")
#max(dups$counts)
#dups[dups$counts==125, ]
dup125<- df[ df$BGN_DATE=="11/11/1995 0:00:00" & df$COUNTY==0 & df$EVTYPE=="HIGH WINDS" , ]
#View(head(dup125))
```
### Now, the duplicates on 5 columns STATEID, BGN_DATE,  COUNTY, EVTYPE

```{r   message=FALSE, warning=FALSE}
#duplicates on 5 variables
library(sqldf)
names(df)[names(df)=="STATE__"]<- "STATEID"
dups<- sqldf(" select STATEID, BGN_DATE,  COUNTY, EVTYPE, count(*) as counts from df group by 1,2,3 having count(*) >1 ")
#max(dups$counts)
#dups[dups$counts==72, ]
dup125<- df[ df$BGN_DATE=="11/11/1995 0:00:00" & df$COUNTY==0 & df$EVTYPE=="HIGH WINDS" & df$STATEID==51 , ]
#View(head(dup125))
```
 
## Adding further self explanatory data cleaning and summarising steps

We carried out 3 main data tidying steps
1. Convert the dates to actual date format, also extract just the year part
2. Convert the damage amounts to actual dollar amounts
3. Sum up the damage and casualties by day, year and finally only on the event type.

```{r   message=FALSE, warning=FALSE}
#Using sqldf as it brings the results in tabular format by default
library(sqldf)
library(lubridate)
# A new date column added for futher processing
df$DATE<- as.Date( as.character( df$BGN_DATE  ) ,"%m/%d/%Y" )

#The property damage and crop damage COSTS need further processing. 
df$CROP_DOL<- as.numeric( 0)
df$PROP_DOL<- as.numeric( 0)
    #Now give them values accordingly, B, M and K 

df[df$CROPDMGEXP %in% c("B","b"),]$CROP_DOL <- df[df$CROPDMGEXP %in% c("B","b"),]$CROPDMG*1000000000
df[df$CROPDMGEXP %in% c("M","m"),]$CROP_DOL <- df[df$CROPDMGEXP %in% c("M","m"),]$CROPDMG*1000000
df[df$CROPDMGEXP %in% c("K","k"),]$CROP_DOL <- df[df$CROPDMGEXP %in% c("K","k"),]$CROPDMG*1000
df[df$CROPDMGEXP %in% c("H","h"),]$CROP_DOL <- df[df$CROPDMGEXP %in% c("H","h"),]$CROPDMG*100
df[ !(df$CROPDMGEXP %in% c("H","h","M","m","K","k","B","b") ),]$CROP_DOL <- df[!(df$CROPDMGEXP %in% c("H","h","M","m","K","k","B","b") ),]$CROPDMG

df[df$PROPDMGEXP %in% c("B","b"),]$PROP_DOL <- df[df$PROPDMGEXP %in% c("B","b"),]$PROPDMG*1000000000
df[df$PROPDMGEXP %in% c("M","m"),]$PROP_DOL <- df[df$PROPDMGEXP %in% c("M","m"),]$PROPDMG*1000000
df[df$PROPDMGEXP %in% c("K","k"),]$PROP_DOL <- df[df$PROPDMGEXP %in% c("K","k"),]$PROPDMG*1000
df[df$PROPDMGEXP %in% c("H","h"),]$PROP_DOL <- df[df$PROPDMGEXP %in% c("H","h"),]$PROPDMG*100
df[ !(df$PROPDMGEXP %in% c("H","h","M","m","K","k","B","b") ),]$PROP_DOL <- df[!(df$PROPDMGEXP %in% c("H","h","M","m","K","k","B","b") ),]$PROPDMG

#Now we can summarise in dollar amount by event type and event date (not by date here)
df$YEAR<- year(df$DATE)
df$EVTYPE<- tolower(trimws(df$EVTYPE, which="both"  ) )
sum_evtype<- sqldf("select EVTYPE
                ,sum(FATALITIES) as FATALITIES
                   ,sum(INJURIES) as INJURIES 
                   , sum(PROP_DOL) as PROP_DOL
                   , sum(CROP_DOL)  as CROP_DOL
                   from df group by 1  ")
rm(df) #remove the big data frame

```

### It seems we have 985 different event types
### Which was not unexpected. Need to do some text mining to 
### put these into the same category

We do the following here
1. Remove invalid event types ( events which are actually Summaries )
2. Build a WordCloud based on the extent of life and property damage separately.

### WordCloud for Loss of Lives and Injuries
===========================================================

```{r   message=FALSE, warning=FALSE}
#1. Remove the rows with 'summary' in them in the EVTYPE column
source("cleancorpus.R")
source("tm_wc.R")
valid_ev<- sum_evtype[ which( !grepl("summary", sum_evtype$EVTYPE) ) , ]
#2.get the character collection in a character vector; formulae is repeatition
charcol<- rep(valid_ev$EVTYPE, valid_ev$FATALITIES +( valid_ev$INJURIES*0.5 )   )
tm_wc(charcol)
```

### WordCloud for Crop and Property Damage
===========================================================

```{r   message=FALSE, warning=FALSE}
#3. get the same for Property and Crop damage
charcol<- rep(valid_ev$EVTYPE, 1/10000000*( valid_ev$CROP_DOL + valid_ev$PROP_DOL)   )
tm_wc(charcol)
```

